Estimating real-time highstreet footfall from Wi-Fi probe requests

### 2019-01-22 stages/second_submission/sections/conclusion.tex
added stages of submission

Sentient technologies make possible measurement of the human activities that are
the life blood of the smart city. Yet the data that they harvest are frequently
relevant only to sub-groups within society that avail themselves of particular
goods and services – such as social media applications, transport modes or
retail offers. In each of these cases, it is necessary to remember that the
resulting data are a by-products of consumer transactions, and will as a
consequence, only pertain to users of the relevant good or service. If the smart
city is to be socially inclusive, it therefore follows that sentient data must
represent entire populations, whether by design or by triangulation with
external, population wide, sources. This usually poses a non-trivial task, since
the ebbs and flows of smart device-enabled citizens rarely pertain to any
clearly defined population in either administrative or functional terms Massam
1973.

Our objective here has been to collect, rather than re-use, data on smart city
functioning, by recording Wi-Fi probes and ultimately reconciling them with
manual counts in order to infer ambient populations. The internal validation
methodology set out in the technical sections of this paper, allied to external
validation from pedestrian counts, renders the method inclusive and robust when
recording activity levels in retail centres in real time. We have described the
collection and processing of a novel consumer Big Dataset that enables valid
measures of levels of footfall activity which has been scaled across a Great
Britain wide network of sensors \citep{cdrc2018consumer}.  In both conceptual and technical
terms, it illustrates the ways in which passively collected consumer data can be
‘hardened’ to render them robust and reliable by using related procedures of
internal and external validation.

Internal validation addresses the issues of screening out device probes that do
not indicate footfall, and the further screening of device probes to
‘fingerprint’ the effects of MAC randomization. It is important to note that the
filtering process work based solely on the information present in the probe
requests and their temporal distribution. This ensures that although the mobile
devices were uniquely identified, there was no further personal data generated
by linking the probe requests to the users of the mobile devices. This method
essentially gave us a way to estimate the footfall in real-time without
identifying or tracking the mobile devices themselves. External validation then
entails reconciling adjusted counts with the footfall observed at sample
locations. This procedure makes it possible to generalise from locations at
which manual footfall surveys are conducted to all others in the system, and to
develop a classification of device locations that are more or less susceptible
to noise generation.

This Wi-Fi based footfall counting methodology offers a large number of
applications and benefits for real time spatial analysis. Since Wi-Fi based
sensors are inexpensive and the data model is scalable, it is possible to use
this methodology for a large network of sensors to gather granular data on
pedestrian footfall. A snapshot showing week’s worth of precise footfall in area
around Charring cross, London is shown in Figure 7 to demonstrate the potential
for such dataset. Projects such as SmartStreetSensors \citep{cdrc2018consumer}, may utilise
this methodology to overcome the challenges introduced by the implementation of
MAC address randomisation.

The vicissitudes of MAC randomisation, and the provisions of privacy legislation
such as EU General Data Protection Regulations mitigate against tracking
individuals across the smart city using this approach.  This can be modelled
using agent-based methods \citep{heppenstall2011agent}, however. In our own research
we have also begun to link store time-lagged till receipts to footfall, and have
used such data to better understand the dwell times that characterise such
different retail uses as stores with window displays and fast food restaurants.
Such analysis not only provides a more nuanced picture of movement through
retail areas, but also enables valorisation of micro sites within retail
centres. In the UK, for example, this is of immediate practical importance in
evaluating business rates on properties, and has still wider implications for
the setting of retail unit rental values. There are obvious extensions to
understanding the ebbs and flows of activities in the 24 hour smart city.

More broadly still, extensions to this strand of smart city research are likely
to seek to differentiate the quality of different elements within footfall
according to mission e.g. travel to adjacent workplace zones, leisure, etc and
personal characteristics such as spending power. In this respect, future
research may not only simulate linkage of harmonised footfall counts between
sensor locations, but also link these in turn to disaggregate origin-destination
matrices for bikeshare and other public transport modes.  Our own investigations
will consider these and other challenges to understanding the functioning of the
sentient city.

>>>>

Sentient technologies make measurement of the human activities that are the life
blood of the smart city possible. Yet the data that they harvest are frequently
relevant only to the sub-groups within society that avail themselves of
particular goods and services – such as social media applications, transport
modes or retail offers. In each of these cases, it is necessary to remember that
the resulting data are by-products of consumer transactions, and will as a
consequence, only pertain to users of the relevant goods or services. If the
smart city is to be socially inclusive, it therefore follows that sentient data
must represent entire populations, whether by design or by triangulation with
external, population wide, sources. This is a non-trivial task, since the ebbs
and flows of smart device-enabled citizens rarely pertain to any clearly defined
population in either administrative or functional terms
\citep{massam1975location}.

Our objective here has been to collect, rather than re-use, data on smart city
functioning, by recording Wi-Fi probes and ultimately reconciling them with
manual counts in order to infer ambient populations. The internal validation
methodology set out in the technical sections of this paper, allied to external
validation from pedestrian counts, renders the method inclusive and robust when
recording activity levels in retail centres in real time. We have described the
collection and processing of a novel consumer Big Dataset that enables valid
measures of levels of footfall activity which has been scaled across a wide
network of sensors \citep{cdrc2018consumer}.  In both conceptual and technical
terms, it illustrates the ways in which passively collected consumer data can be
‘hardened’ to render them robust and reliable by using related procedures of
internal and external validation.

Internal validation addresses the issues of screening out device probes that do
not indicate footfall, and the further screening of device probes to
‘fingerprint’ the effects of MAC randomization. It is important to note that the
filtering process work based solely on the information present in the probe
requests and their temporal distribution. This ensures that although the mobile
devices were uniquely identified, there was no further personal data generated
by linking the probe requests to the users of the mobile devices. This method
essentially gave us a way to estimate the footfall in real-time without
identifying or tracking the mobile devices themselves. External validation then
entailed reconciling adjusted counts with the footfall observed at sample
locations. This procedure makes it possible to generalise from locations at
which manual footfall surveys are conducted to all others in the system, and to
develop a classification of device locations that are more or less susceptible
to noise generation.

This Wi-Fi based footfall counting methodology offers a large number of
applications and benefits for real time spatial analysis. Since Wi-Fi based
sensors are inexpensive and the data model is scalable, it is possible to use
this methodology for a large network of sensors to gather granular data on
pedestrian footfall. A snapshot showing a week’s worth of precise footfall in
area around Charring cross, London is shown in Figure \ref{main_study_counts} in
order to demonstrate the potential for such a dataset. Projects such as
SmartStreetSensors \citep{cdrc2018consumer}, may utilise this methodology to
overcome the challenges introduced by the implementation of MAC address
randomisation.

The vicissitudes of MAC randomisation, and the provisions of privacy legislation
such as EU General Data Protection Regulations mitigate against tracking
individuals across the smart city using this approach.  This can be modelled
using agent-based methods \citep{heppenstall2011agent}, however. In our own
research we have also begun to link store time-lagged till receipts to footfall,
and have used such data to better understand the dwell times that characterise
such different retail uses as stores with window displays and fast food
restaurants.  Such analysis not only provides a more nuanced picture of movement
through retail areas, but also enables valorisation of micro sites within retail
centres. In the UK, for example, this is of immediate practical importance in
evaluating business rates on properties, and has still wider implications for
the setting of retail unit rental values. There are obvious extensions to
understanding the ebbs and flows of activities in the 24-hour smart city.

More broadly still, extensions to this strand of smart city research are likely
to seek to differentiate the quality of different elements within footfall
according to mission e.g. travel to adjacent workplace zones, leisure, etc., and
personal characteristics such as spending power. In this respect, future
research may not only simulate linkage of harmonised footfall counts between
sensor locations, but also link these in turn to disaggregate origin-destination
matrices for bikeshare and other public transport modes.  Our own investigations
will consider these and other challenges to understanding the functioning of the
sentient city.

### 2019-01-22 stages/second_submission/sections/main_study.tex
added stages of submission

\begin{table}
    \tbl{Locations where sensors were installes, volume and speed of probe requests collected by
    the sendor and total pedestrians manually counted. The data occupies around 1.8 GB on disk 
    when encoded in text format.}
	{\begin{tabular}{clllccc} 
		\toprule
         ID & Location & Type & Installation notes & Probe Requests & Footfall\\
         & & & & x10\textsuperscript{6} (per min) & No. (per min)\\
		 \midrule
         1 & Camden High Street & Phone Shop & Bus stop in front & 9.9 (297) & 3683 (33)\\
         2 & Central St.Giles & Restaurant & Seating area on both sides & 3.9 (169) & 0346 (05)\\
         3 & Holborn Station & Info. Kiosk & Overlooks station entrance & 4.3 (303) & 2956 (46)\\
         4 & Brunswick Center & Fast Food & Has seating area on one side & 3.4 (210) & 0960 (12)\\
         5 & The Strand & Tea Shop & Has phone shop next door & 8.4 (382) & 1969 (21)\\
		 \bottomrule
	\end{tabular}}
	\label{locations-table}
\end{table}

>>>>

\begin{table}
	\tbl{Locations where sensors were installed}
	{\begin{tabular}{clll} 
		\toprule
		 ID & Location & Type & Installation notes\\
		 \midrule
		 1 & Camden High Street & Phone Shop & Bus stop in front\\
		 2 & Central St.Giles Piazza & Restaurant & Seating area on both sides\\
		 3 & Holborn Underground Station & Information Kiosk & Overlooks station entrance\\
		 4 & Brunswick Center & Fast Food Restaurant & Has seating area on one side\\
		 5 & The Strand & Tea Shop & Has phone shop next door \\
		 \bottomrule
	\end{tabular}}
	\label{locations-table}
\end{table}

### 2019-01-22 stages/second_submission/sections/methodology.tex
added stages of submission

The nodes were then assigned a unique ID based on the unique connected component
they belonged to as shown in Figure \ref{methodology_schematic}. This unique
identifier was used in the place of MAC addresses for aggregation of the
anonymised probe requests. Although the recycling of sequence numbers after 4096
led to multiple unique IDs being reported from a single device, a sample
consisting of all randomised probe requests sent by "Google" devices showed that
only 0.05\% of the sample had their sequence number reset. This led assume this
to be inconsequential.

>>>>

The nodes were then assigned a unique ID based on the unique connected component
they belonged to as shown in Figure \ref{methodology_schematic}. This unique
identifier was used in the place of MAC addresses for aggregation of the
anonymised probe requests. Although the recycling of sequence numbers after 4096
led to multiple unique IDs being reported from a single device, a sample
consisting of all randomised probe requests sent by "Google" devices showed that
only 0.5\% of the sample had their sequence number reset. This led assume this
to be inconsequential.

### 2019-01-22 stages/second_submission/sections/pilot_study.tex
added stages of submission

We then classified the probe requests as `high signal strength' and `low signal
strength' using various one dimensional clustering algorithms such as k-means,
quantile, hierarchical clustering, bagged clustering, fisher and jenks natural
breaks with the number of clusters set as 2. The results are shown in Table
\ref{classification-table}. We found that while hierarchical clustering and
jenks gave us fairly low errors, they were too resource intensive for practical
use with a larger dataset. We also found that k-means gave the quickest results
with the lowest MAPE, closely followed by quantile algorithm. The cut-off point
or threshold for the collected data with which we could classify as high and
low was -71 dBm. We then removed all the probe requests which reported `low
signal strength' and repeated the same aggregation process as before to produce
footfall count. This process resulted in a footfall count with a net MAPE of
30\%. Although the results are encouraging we are still not completely
confident that our filtering process is removing noise or has any correlation
the configuration of sensor or position of the mobile devices. These concerns
need to be addressed with a larger survey with multiple locations of varying
orientations.

\begin{figure}
\begin{center}
\vspace{10pt}
\includegraphics [width=0.85\linewidth,trim=10 15 10 15,clip]
    {images/pilot_clustering_params_rev.png}
\caption{The clustering process was repeated with both increasing sequence
    number threshold ($\alpha$) and time threshold ($\beta$), until we arrived
    at the lowest parameters where the know device (black line) is clustered as
    a single device.}
\label{pilot_clustering_params}
\end{center}
\end{figure}

\begin{figure}
\centering
\subfigure[Clustering probe requests based on increasing sequence numbers 
	present in them.]
	{\resizebox*{0.45\linewidth}{!}
	{\includegraphics{images/pilot_clustering_rev1.jpeg}}}
\hspace{20pt}
\subfigure[Comparison of sensor based counts with the ground truth.]
	{\resizebox*{0.45\linewidth}{!}
	{\includegraphics{images/pilot_comparison.jpeg}}}
\caption{The results of the pilot study demonstrating the validity of the
	methodology.} 
\label{pilot_results}
\end{figure}

The next challenge was to identify the probe requests which are generated by
the same device irrespective of the MAC randomisation process. We used the
algorithm defined in Section \ref{methodology} and assigned a unique identifier
or signature to each probe request, independent of their the MAC addresses.
Since we didn't know the nature or frequency of the MAC address randomisation
process, we used the surveyor's mobile device as a reference. As the surveyor's
device was being actively used to count pedestrians and it's Wi-Fi module was
kept active without establishing connection to any network, the device was
known to be continuously probing for new networks. We also knew that the OUI of
the device was 'Google' and the device was regularly randomising it's MAC
address, thus providing us an excellent reference with which we could optimise
the parameters for our clustering algorithm. We then, by increasing the
thresholds in steps of 1, found the minimum possible threshold for time and
sequence numbers at which the algorithm clusters the reference device properly
without over clustering the other probe requests.  This process is shown in
Figure \ref{pilot_clustering_params}. We observed that the threshold for time
$\alpha$ and the threshold for sequence numbers, $\beta$ are 16 seconds and 60
respectively This was undertaken on top the filtering done based on signal
strength, and only for the probe requests with randomised MAC addresses. Figure
\ref{pilot_results} shows the results of this clustering process on a small set
of randomised probe requests. The probe requests with different randomised MAC
address are shown by the coloured points and the lines joining them show that
those probe requests were most likely be generated by the same device. We
finally aggregated the probe requests as we did before but with the device
signature rather than MAC addresses. This results in a footfall count with a
MAPE of -18\% compared to the manual count. A comparison of minute by minute
counts resulting from different filtering processes along with the ground truth
is shown in Figure \ref{pilot_results} illustrating the promising effectiveness
of the methods.

>>>>

We then classified the probe requests as `high signal strength' and `low signal
strength' using various one dimensional clustering algorithms such as k-means,
quantile, hierarchical clustering, bagged clustering, fisher and jenks natural
breaks. The results are shown in Table \ref{classification-table}. We found that
while hierarchical clustering and jenks gave us fairly low errors, they were too
resource intensive for practical use with a larger dataset. We also found that
k-means gave the quickest results with the lowest MAPE, closely followed by
quantile algorithm. The cut-off point or threshold for the collected data with
which we could classify as high and low was -71 dBm. We then removed all the
probe requests which reported `low signal strength' and repeated the same
aggregation process as before to produce footfall count. This process resulted
in a footfall count with a net MAPE of 30\%. Although the results are
encouraging we are still not completely confident that our filtering process is
removing noise or has any correlation the configuration of sensor or position of
the mobile devices. These concerns need to be addressed with a larger survey
with multiple locations of varying orientations.

\begin{figure}
\begin{center}
\vspace{10pt}
\includegraphics [width=\linewidth,trim=10 30 10 75,clip]
    {images/pilot_clustering_params.png}
\caption{The clustering process was repeated with both increasing sequence
    number threshold ($\alpha$) and time threshold ($\beta$), until we arrived
    at the lowest parameters where the know device (black line) is clustered as
    a single device.}
\label{pilot_clustering_params}
\end{center}
\end{figure}

\begin{figure}
\centering
\subfigure[Clustering probe requests based on increasing sequence numbers 
	present in them.]
	{\resizebox*{0.45\linewidth}{!}
	{\includegraphics{images/pilot_clustering.jpeg}}}
\hspace{20pt}
\subfigure[Comparison of sensor based counts with the ground truth.]
	{\resizebox*{0.45\linewidth}{!}
	{\includegraphics{images/pilot_comparison.jpeg}}}
\caption{The results of the pilot study demonstrating the validity of the
	methodology.} 
\label{pilot_results}
\end{figure}

The next challenge was to identify the probe requests which are generated by the
same device irrespective of the MAC randomisation process. We used the algorithm
defined in Section \ref{methodology} and assigned a unique identifier or
signature to each probe request, independent of their the MAC addresses. Since
we didn't know the nature or frequency of the MAC address randomisation process,
we used the surveyor's mobile device as a reference. As the surveyor's device
was being actively used to count pedestrians and it's Wi-Fi module was kept
active without establishing connection to any network, the device was known to
be continuously probing for new networks. We also knew that the OUI of the
device was 'Google' and the device was regularly randomising it's MAC address,
thus providing us an excellent reference with which we could optimise the
parameters for our clustering algorithm. Using this reference device we observed
that the threshold for time $\alpha$ and the threshold for sequence numbers,
$\beta$ are 16 seconds and 60 respectively via trial and error. This process is
shown in Figure \ref{pilot_clustering_params}. This was undertaken on top the
filtering done based on signal strength, and only for the probe requests with
randomised MAC addresses. Figure \ref{pilot_results} shows the results of this
clustering process on a small set of randomised probe requests. The probe
requests with different randomised MAC address are shown by the coloured points
and the lines joining them show that those probe requests were most likely be
generated by the same device. We finally aggregated the probe requests as we did
before but with the device signature rather than MAC addresses. This results in
a footfall count with a MAPE of -18\% compared to the manual count. A comparison
of minute by minute counts resulting from different filtering processes along
with the ground truth is shown in Figure \ref{pilot_results} illustrating the
promising effectiveness of the methods.

### 2019-01-22 stages/third_submission/sections/conclusion.tex
added stages of submission

The vicissitudes of MAC randomisation, and the provisions of privacy legislation
such as EU General Data Protection Regulations mitigate against tracking
individuals across the smart city using this approach.  This can be modelled
using agent-based methods \citep{heppenstall2011agent}, however. In our own
research we have also begun to link store time-lagged till receipts to footfall,
and have used such data to better understand the dwell times that characterise
such different retail uses as stores with window displays and fast food
restaurants.  Such analysis not only provides a more nuanced picture of movement
through retail areas, but also enables valorisation of micro sites within retail
centres. In the UK, for example, this is of immediate practical importance in
evaluating business rates on properties, and has still wider implications for
the setting of retail unit rental values. There are obvious extensions to
understanding the ebbs and flows of activities in the 24-hour smart city.

>>>>

The vicissitudes of MAC randomisation, and the provisions of privacy
legislation such as EU General Data Protection Regulations mitigate against
tracking individuals across the smart city using this approach.  This can be
modelled using agent-based methods \citep{heppenstall2011agent}, however. In
our own research we have also begun to link store time-lagged till receipts to
footfall, and have used such data to better understand the dwell times that
characterise such different retail uses as stores with window displays and fast
food restaurants.  Such analysis not only provides a more nuanced picture of
movement through retail areas, but also enables valorisation of micro sites
within retail centres. In the UK, for example, this is of immediate practical
importance in evaluating business rates on properties, and has still wider
implications for the setting of retail unit rental values. There are obvious
extensions to understanding the ebbs and flows of activities in the 24-hour
smart city such as understanding urban mobility \citep{gariazzo2019spatio} and
conceptualising them with a people dimension \citep{nam2011conceptualizing}.

### 2019-01-22 stages/third_submission/sections/previous_work.tex
added stages of submission

In the past decade, Wi-Fi has emerged as one of the most commonly used
technologies in providing high speed internet access to mobile devices such as
smartphones, tablets and laptops in public and private spaces. This has
resulted in multiple Wi-Fi networks being available at almost every location in
dense urban environments. Traversing through this overlapping mesh of Wi-Fi
networks, modern mobile devices with Wi-Fi network interfaces regularly
broadcast a special type of signal known as `Probe Requests' in order to
discover the Wi-Fi networks available to them. This helps these devices to
connect and switch between the Wi-Fi networks seamlessly.

@@

Unlike GPS, the location of the Wi-Fi enabled mobile device cannot be directly
inferred from Wi-Fi, however there are reliable methods to triangulate the
location of mobile devices from the locations of known access points (AP) and
the signal strength reported by them \citep{he2003range, moore2004robust,
lamarca2005place}. This can overcome the usual shortcoming of GPS, which
struggles for precision and accuracy in indoor and densely built environments
\citep{zarim2006, kawaguchi2009wifi, xi2010locating}. Utilising this, we can
easily and quickly estimate trajectories of the mobile devices
\citep{Sorensenlocation, musa2012tracking} which can be used similary to the
GPS trajectories to understand individual travel patterns \citep{reki2007,
Sap2015}, crowd behaviour \citep{abedi2013bluetooth, mowafi2013tracking},
vehicular \citep{lu2010vehicle} and pedestrian movement
\citep{xu2013pedestrian, fukuzaki2014pedestrian, wang2016gait}. Such data can
also be used in transportation planning and management to estimate travel time
\citep{musa2011wiflow} and real time traffic monitoring
\citep{abbott2013empirical}. Using techniques demonstrated by
\cite{franklin2006passive} and \cite{pang2007802}, along with information
present in the probe requests, one can even model interactions between the
users \citep{cheng2012inferring, barbera2013signals, cunche2014know,
cunche2014linking} such as predicting which of them are most likely to meet
again \citep{cunche2012know}. Using the semantic information present in these
probe requests it even is possible to understand the nature of population at a
large scale \citep{di2016mind}.

>>>>

In the past decade, Wi-Fi has emerged as one of the most commonly used
technologies in providing high speed internet access to mobile devices such as
smartphones, tablets and laptops in public and private spaces
\citep{torrens2008wi}. This has resulted in multiple Wi-Fi networks being
available at almost every location in dense urban environments. Traversing
through this overlapping mesh of Wi-Fi networks, modern mobile devices with
Wi-Fi network interfaces regularly broadcast a special type of signal known as
`Probe Requests' in order to discover the Wi-Fi networks available to them.
This helps these devices to connect and switch between the Wi-Fi networks
seamlessly.

@@

Unlike GPS, the location of the Wi-Fi enabled mobile device cannot be directly
inferred from Wi-Fi, however there are reliable methods to triangulate the
location of mobile devices from the locations of known access points (AP) and
the signal strength reported by them \citep{he2003range, moore2004robust,
lamarca2005place}. This can overcome the usual shortcoming of GPS, which
struggles for precision and accuracy in indoor and densely built environments
\citep{zarim2006, kawaguchi2009wifi, xi2010locating}. Utilising this, we can
easily and quickly estimate trajectories of the mobile devices
\citep{musa2012tracking} which can be used similary to the
GPS trajectories to understand individual travel patterns \citep{reki2007,
Sap2015}, crowd behaviour \citep{abedi2013bluetooth, mowafi2013tracking},
vehicular \citep{lu2010vehicle} and pedestrian movement
\citep{xu2013pedestrian, fukuzaki2014pedestrian, wang2016gait}. Such data can
also be used in transportation planning and management to estimate travel time
\citep{musa2011wiflow} and real time traffic monitoring
\citep{abbott2013empirical}. Using techniques demonstrated by
\cite{franklin2006passive} and \cite{pang2007802}, along with information
present in the probe requests, one can even model interactions between the
users \citep{cheng2012inferring, barbera2013signals, cunche2014know,
cunche2014linking} such as predicting which of them are most likely to meet
again \citep{cunche2012know}. Using the semantic information present in these
probe requests it even is possible to understand the nature of population at a
large scale \citep{di2016mind}.
